Numerical simulations of the Cordilleran ice sheet through the last glacial cycle

### 2015-06-20 cordillera-cycle.tex
Add corrections from Arjen and Martin (part 2).

\begin{figure}
  \includegraphics{locmap}
  \caption{Relief map of northern North America showing a reconstruction of the
           areas once covered by the Cordilleran (CIS), Laurentide (LIS),
           Innuitian (IIS), and Greenland (GIS) ice sheets during the last
           18\,\unit{\chem{^{14}C}\,ka} \citep[21.4\,cal\,ka,][]{Dyke.2004}.
           The rectangular box denotes the location of the
           modelling domain used in this study. Major mountain ranges covered
           by the ice sheet include the Alaska Range (AR), the Wrangell and
           St.-Elias mountains (WSEM), the Selwyn and Mackenzie mountains
           (SMKM), the Skeena Mountains (SM), the Coast Mountains (CM), the
           Columbia and Rocky
           Mountains (CRM), and the North Cascades (NC). Major depressions
           include the Liard Lowland (LL), the Interior Plateau of British
           Columbia (IP), and the Puget Lowland (PL). The background
           map consists of ETOPO1 \citep{Amante.Eakins.2009} and Natural Earth
           Data \citep{Patterson.Kelso.2015}.}
  \label{fig:locmap}
\end{figure}

@@

A~pseudo-plastic sliding law,
\begin{equation}
    \label{eqn:pseudoplastic}
    \vec{\tau}_b = -\tau_c \frac{\vec{v}_b}{{v_{th}}^q\,|\vec{v}_b|^{1-q}} \,,
\end{equation}
relates the bed-parallel shear stresses, $\vec{\tau}_b$, to the sliding
velocity, $\vec{v}_b$ (Table~\ref{tab:params}).
The yield stress, $\tau_c$,
is modelled using the Mohr--Coulomb criterion,
\begin{equation}
   \tau_c = c_0 + N\,\tan{\phi} \,,
\end{equation}
where cohesion, $c_0$, is assumed to be zero. The friction angle, $\phi$,
varies from 15 to 45\degree as a piecewise-linear function of modern bed
elevation, with the lowest value occuring below modern sea level (0\,m above
sea level, m~a.s.l.) and the highest value occuring above the generalised
elevation of the highest shorelines
\citep[200\,m~a.s.l.,][Fig.~5]{Clague.1981}, thus accounting for
a~weakening of till associated with the presence of marine sediments
(cf. \citealp{Martin.etal.2011}; \citealp[supplement]{Aschwanden.etal.2013};
\citealp{PISM-authors.2015}). Effective pressure, $N$,
is related to the ice overburden stress, $\rho gh$, and the modelled amount of
subglacial water, using a formula derived from laboratory experiments with till
extracted from the base of Ice Stream B in West Antarctica
\citep{Tulaczyk.etal.2000, Bueler.Pelt.2014},
\begin{equation}
    N = \delta \rho gh \, 10^{(e_0/C_c) (1 - (W/W_{max}))} \,,
\end{equation}
where $\delta$ is chosen as 0.02, $e_0$ is a measured reference void ratio and
$C_c$ is a measured compressibility coefficient (Table~\ref{tab:params}). The
amount of water at the base, $W$, varies from zero to $W_{max}=2$\,m, a
threshold above which additional melt water is assumed to drain off
instantaneously. Finally, the bedrock topography responds to ice load
following a bedrock deformation model that includes local isostasy,
elastic lithosphere flexure and viscous astenosphere deformation in an infinite
half-space \citep{Lingle.Clark.1985,Bueler.etal.2007}.
A relatively low viscosity value of $\nu_m = 1\times10^{19}$\,\unit{Pa\,s} is
used for the astenosphere (Table~\ref{tab:params}) in accordance with the
results from regional glacial
isostatic adjustment modelling at the northern Cascadia subduction zone
\citep{James.etal.2009}.

@@

It is generally believed that the Cordilleran ice sheet formed by the
coalescence of several mountain-centred ice caps \citep{Davis.Mathews.1944}.
In our simulations, major ice-dispersal centres, visible on the modelled ice
cover duration maps (Fig.~\ref{fig:duration}), are located over the Coast
Mountains (CM), the Columbia and Rocky mountains (CRM), the Skeena Mountains
(SM), and the Selwyn and Mackenzie mountains (SMKM). The Wrangell and
St.-Elias mountains, heavily glacierized at present, host an ice cap for the
entire length of both simulations, but that ice cap does not appear to be a
major feed to the Cordilleran ice sheet (Fig.~\ref{fig:duration}, WSEM).
Although the Coast, Skeena and Columbia and Rocky mountains (CM, SM, CRM)
are covered by mountain
glaciers for most of the last glacial cycle, providing durable nucleation
centres for an ice sheet, this is not the case for the Selwyn and
Mackenzie mountains (SMKM), where ice cover on the highest peaks is limited to a
small fraction of the last glacial cycle. In other words, the Selwyn and Mackenzie
mountains only appear as a secondary ice-dispersal centre
during the coldest periods of the last glacial cycle. The Northern Rocky
Mountains (Fig.~\ref{fig:duration}, NRM) do not act as a nucleation centre,
but rather as a pinning point for the Cordilleran ice sheet margin coming from
the west.

@@

\begin{figure*}
  \includegraphics{warmbase}
  \caption{Modelled duration of warm-based ice cover during the last
           120\,ka. Long ice cover durations combined with basal
           temperatures at the pressure-melting point may explain the strong
           glacial erosional imprint of the Skeena Mountains (SM) landscape.
           Hatches indicate areas that were covered by cold ice only.}
  \label{fig:warmbase}
\end{figure*}

@@

A correlation is observed between the modelled duration of warm based ice cover
(Fig.~\ref{fig:warmbase}) and the degree of glacial modification of the
landscape (mainly in terms of the development of deep glacial valleys and
troughs). We find evidence for this on both the southwestern and northeastern
slopes of the southern Coast Mountains, on the western slopes of the Columbia
Mountains, on the western and eastern slopes of the northern Coast Mountains
and the Saint-Elias Mountains, and radiating off the Skeena Mountains
(Figs.~\ref{fig:duration} and \ref{fig:warmbase};
\citealp[Fig.~2]{Kleman.etal.2010}).
The Skeena Mountains, for example, indeed bear a strong glacial imprint that
indicates ice drainage in a system of distinct glacial troughs emanating in a
radial pattern from the centre of the mountain range
\citep[Fig.~2]{Kleman.etal.2010}. We suggest that
persistent ice cover (Fig.~\ref{fig:duration}) associated with basal ice
temperatures at the pressure-melting point (Figs.~\ref{fig:warmbase}
and~\ref{fig:warmfrac}) explains the large-scale glacial erosional imprint on
the landscape. A well-developed network of glacial valleys west of
the Mackenzie Mountains (\citealp[Fig.~2]{Kleman.etal.2010}; \citealp[Fig.~8]
{Stroeven.etal.2010}) is modelled to have hosted mainly warm-based ice
(Fig.~\ref{fig:warmfrac}). However, because it is only glaciated for a
short fraction of the last glacial cycle in our simulations
(Fig.~\ref{fig:duration}), this perhaps indicates that
the observed landscape pattern originates from multiple glacial cycles and
witnesses an increased relative importance of the Selwyn and Mackenzie
mountains ice dispersal centre (Fig.~\ref{fig:duration}, SMKM), prior
to the Late Pleistocene \citep[cf.][]{Ward.etal.2008, Demuro.etal.2012}.

@@

Similarly to other glaciated regions, most glacial traces in the North American
Cordillera relate to the last few millennia of glaciation, because most of the
older evidence has been overprinted by warm-based ice retreat
during the last deglaciation \citep{Kleman.1994, Kleman.etal.2010}. From a
numerical modelling perspective, phases of glacier retreat are more challenging
than phases of growth, because they involve more rapid fluctuations of the ice
margin, increased flow velocities and longitudinal stress gradients, and poorly
understood hydrological processes. The latter are typically included in the
models through simple parametrisations \citep[e.g.][]{Clason.etal.2012,
Clason.etal.2014, Bueler.Pelt.2014}, if included at all. However, next after
the mapping of maximum ice sheet extents during MIS~2 and MIS~4
(Sect.~\ref{sec:mis2} and~\ref{sec:mis4}), geomorphologically-based
reconstructions of patterns of ice sheet retreat during the last
deglaciation provide the second best source of evidence for the validation of
our simulations.

@@

\begin{figure*}
  \includegraphics{deglacshots}
  \caption{Snapshots of modelled surface topography (200\,m contours)
           and surface velocity (colour mapping) during the last deglaciation
           from the GRIP (top panels) and EPICA (bottom panels) 5\,km simulations.
           Dashed segments indicate the location of profiles used in
           Figs.~\ref{fig:profiles-grip} and~\ref{fig:profiles-epica}.}
  \label{fig:deglacshots}
\end{figure*}

\begin{figure*}
  \vspace{-1.5mm}  % enough to allow the two figures to fit on one page
  \includegraphics{deglac}
  \caption{Modelled age of the last deglaciation. Areas where the MIS~4 glacial
           advance exceeded the last glacial maximum advanced are marked in
           green. Hatches denote re-advance of mountain-centred ice caps and
           the decaying ice sheet between 14 and 10\,ka., which is very
           restricted in the EPICA-driven simulation.
           Dashed segments indicate the location of profiles used in
           Figs.~\ref{fig:profiles-grip} and~\ref{fig:profiles-epica}.}
  \label{fig:deglac}
\end{figure*}

@@

Modelled patterns of ice sheet retreat are relatively consistent between the
two simulations (Figs.~\ref{fig:deglacshots} and.~\ref{fig:deglac}). The
southern sector of the modelling domain, including the Puget Lowland, the Coast
and Rocky mountains, and the Interior Plateau of British Columbia, becomes
completely deglaciated by 10\,ka, whereas a significant ice cover remains over
the Skeena, Selwyn, Mackenzie, Wrangell and Saint-Elias mountains in the
northern sector of the modelling domain. After 10\,ka, deglaciation continues
to proceed across the Liard Lowland with a radial ice margin retreat towards
the surrounding mountain ranges, consistent with the regional melt water record
of the last deglaciation \citep{Margold.etal.2013}. Remaining ice continues to
decay by retreating towards the Selwyn and Skeena mountains. The last remnants
of the Cordilleran ice sheet finally disappear from the Skeena Mountains at
6.5\,ka (GRIP) and 6.1\,ka (EPICA).

@@

The possibility of late glacial readvances in the North American Cordillera has
been debated for some time \citep{Osborn.Gerloff.1997}, and locally these have
been reconstructed and dated. Radiocarbon-dated end moraines in the Fraser and
Squamish valleys, off the southern tip of the Coast Mountains, indicate
consecutive glacier maxima, or standstills while in overall retreat, one of
which corresponds to the Younger Dryas chronozone \citep{Clague.etal.1997,
Friele.Clague.2002, Friele.Clague.2002a, Kovanen.2002,
Kovanen.Easterbrook.2002}. Although most of these moraines characterise
independent valley glaciers, that may have been disconnected from the waning
Cordilleran ice sheet, the Finlay River area in the Omenica Mountains
(Fig.~\ref{fig:deglac}, OM) presents a different kind of evidence.
There, sharp-crested moraines indicate a late-glacial readvance of local alpine
glaciers and, more importantly, their interaction with larger, lingering
remnants of the main body of the Cordilleran ice sheet in the valleys
\citep{Lakeman.etal.2008}. Additional evidence for late-glacial
alpine glacier readvances includes moraines in the eastern Coast Mountains,
Rocky Mountains and the Columbia Mountains \citep{Osborn.Gerloff.1997,
Menounos.etal.2008}.

Although further work is needed to constrain the timing of the late-glacial
readvance, to assess its extents and geographical distribution, and to identify
the potential climatic triggers \citep{Menounos.etal.2008}, it is interesting
to note that the simulation driven by the GRIP record produces a late-glacial
readvance in the Coast Mountains and Rocky Mountains (Fig.~\ref{fig:deglac},
left panel). In addition to matching the location of some local readvances, the
GRIP-driven simulation shows that a large remnant of the decaying ice sheet
may still have existed at the time of this late-glacial readvance.
In contrast, the EPICA-driven simulation produces a nearly-continuous
deglaciation with only a tightly restricted late-glacial readvance on the
western slopes of the Saint-Elias and Coast mountains
(Fig.~\ref{fig:deglac}, right panel).

@@

\begin{figure*}
  \includegraphics{lastflow}
  \caption{Modelled deglacial basal ice velocities. Hatches
           indicate areas that remain non-sliding throughout deglaciation
           (22.0--8.0\,ka), notably including parts of the Interior Plateau (IP).
           Note the concentric patterns of deglacial flow in the Liard
           Lowland (LL).
           Sliding grid cells were distinguished from non-sliding grid cells
           using a velocity threshold of 1\,\unit{m\,yr^{-1}}.
           Dashed segments indicate the location of profiles used in
           Figs.~\ref{fig:profiles-grip} and~\ref{fig:profiles-epica}.}
  \label{fig:lastflow}
\end{figure*}

@@

On the Interior Plateau of south-central British Columbia, both simulations
produce a retreat of the ice margin towards the north-east
(Fig.~\ref{fig:deglac}), a pattern which is validated by the geomorphological
and stratigraphical record for ancient pro-glacial lakes dammed by the
retreating ice sheet \citep{Perkins.Brennand.2014}. However, the two
simulations differ in the mode of retreat. The GRIP simulation yields an active
retreat with basal sliding towards the ice margin to the south, whereas the
EPICA simulation produces negligible basal sliding on the plateau during deglaciation
(Fig.~\ref{fig:lastflow}). Yet, the Interior Plateau also hosts an impressive
set of glacial lineations which indicate a substantial eastwards flow component
of the Cordilleran ice sheet \citep{Prest.etal.1968, Kleman.etal.2010}.
The Fraser lineation set could thus present a smoking gun for the
reliability of the presented model results. One explanation for the incongruent
results could therefore be that missing feedback mechanisms between ice sheet
topography and regional climate position the modelled ice divide of the LGM ice
sheet too far to the east (Sect.~\ref{sec:mis2}; Fig.~\ref{fig:mis2};
\citealp{Seguinot.etal.2014}). A more westerly-located LGM ice divide would
certainly result in a different deglacial flow pattern over the Interior
Plateau. However, a more westerly-positioned LGM ice divide would certainly be
associated with a thinner ice sheet than that modelled here. Decreased ice
thickness would not promote warm-based conditions but, on the
opposite, enlarge the region of negligible basal sliding
(Fig.~\ref{fig:lastflow}). Thus, a second explanation for the incongruent
results could be that the Interior Plateau lineation system observed in the
geomorphology predates the deglaciation, as perhaps indicated by some eskers
that appear incompatible with these glacial lineations
\citep[Fig.~9]{Margold.etal.2013a}. Finally, a third explanation could be that
local geothermal heat associated with volcanic activity on the Interior Plateau
could have triggered the basal sliding
\citep[cf. Greenland ice sheet,][]{Fahnestock.etal.2001}.

@@

Numerical simulations of the Cordilleran ice sheet through the last glacial
cycle presented in this study consistently produce two glacial maxima during
MIS~4 (61.9--56.5\,ka, 3.8--8.8\,m~s.l.e.) and MIS~2 (23.2--16.8\,ka,
8.40--8.91\,m~s.l.e.), two periods corresponding to documented extensive
glaciations. This result is independent of the palaeo-temperature record used
among the six selected for this study, and thus can be regarded as a robust model
output, which broadly matches geological evidence. However, the timing of the
two glaciation peaks depends sensitively on which climate record is used to
drive the model. The timing of the LGM is best reproduced by the EPICA and
Vostok Antarctic ice core records. It occurs about 2\,ka too early in the
simulation forced by the GRIP ice core record, and even earlier in all other
simulations. The mismatch is largest for the two Northwest Pacific ODP
palaeo-temperature records, which are affected by the weakening of the
California current during the LGM.

In all simulations presented here, ice cover is limited to disjoint mountain
ice caps during most of the glacial cycle. The most persistent nucleation
centres are located in the Coast Mountains, the Columbia and Rocky mountains,
the Selwyn and Mackenzie mountains, and most importantly, in the Skeena
Mountains. Throughout the modelled last glacial cycle, the Skeena Mountains
host an ice cap which appears to be fed by the moisture intruding inland from the
west through a topographic breach in the Coast Mountains. The Skeena ice cap acts
as the main nucleation centre at the onset of the pre-LGM readvance. As indicated
by persistent, warm-based ice in the model, this ice cap perhaps explains
the distinct glacial erosional imprint observed on the landscape of the Skeena
Mountains.

During deglaciation, none of the climate records used can be selected as
producing an optimal agreement between the model results and the geological
evidence. Although the EPICA-driven simulation yields the most realistic timing
of the LGM and, therefore, start of deglaciation, only the GRIP-driven
simulation produces late glacial readvances in areas where these have been
documented. Nonetheless, the patterns of ice sheet retreat are consistent
between the two simulations, and show a rapid deglaciation of the southern
sector of the ice sheet, including a rapid northwards retreat across the Interior
Plateau of central British Columbia. The GRIP-driven simulation then produces a
late-glacial readvance of local ice caps and of the main body of the decaying
Cordilleran ice sheet primarily in the Coast, Columbia and Rocky Mountains.
In both simulations, this is followed by an
opening of the Liard Lowland, and a final retreat of the remaining ice caps
towards the Selwyn and, finally, the Skeena mountains, which hosts the last
remnant of the ice sheet during the middle Holocene (6.6--6.2\,ka). Our results
identify the Skeena Mountains as a key area to understanding glacial dynamics
of the Cordilleran ice sheet, highlighting the need for further geological
investigation of this region.

@@

% Acknowledgements
\begin{acknowledgements}
Foremost, we would like to thank Shawn Marshall for providing a detailed,
constructive analysis of this study during J.~Seguinot's Ph.D. defence
(September 2014). His comments were used to improve the model set-up.
We are very thankful to Constantine Khroulev, Ed Bueler and Andy Aschwanden for
providing constant help and development with PISM. This work was supported by the
Swedish Research Council~(VR) grant no.~2008-3449 to A.~P.~Stroeven and by the
German Academic Exchange Service~(DAAD) grant no.~50015537 and a Knut and Alice
Wallenberg Foundation grant to J.~Seguinot. Computer resources were provided
by the Swedish National Infrastructure for Computing (SNIC) allocation
no.~2013/1-159 and~2014/1-159 to A.~P.~Stroeven at the National Supercomputing
Center (NSC).
\end{acknowledgements}

>>>>

\begin{figure}
  \includegraphics{locmap}
  \caption{Relief map of northern North America showing a reconstruction of the
           areas once covered by the Cordilleran (CIS), Laurentide (LIS),
           Innuitian (IIS), and Greenland (GIS) ice sheets during the last
           18\,\unit{\chem{^{14}C}\,ka} \citep[21.4\,cal\,ka,][]{Dyke.2004}.
           The rectangular box denotes the location of the
           modelling domain used in this study. Major mountain ranges covered
           by the ice sheet include the Alaska Range (AR), the Wrangell and
           Saint Elias mountains (WSEM), the Selwyn and Mackenzie mountains
           (SMKM), the Skeena Mountains (SM), the Coast Mountains (CM), the
           Columbia and Rocky
           Mountains (CRM), and the North Cascades (NC). Major depressions
           include the Liard Lowland (LL), the Interior Plateau of British
           Columbia (IP), and the Puget Lowland (PL). The background
           map consists of ETOPO1 \citep{Amante.Eakins.2009} and Natural Earth
           Data \citep{Patterson.Kelso.2015}.}
  \label{fig:locmap}
\end{figure}

@@

A~pseudo-plastic sliding law,
\begin{equation}
    \label{eqn:pseudoplastic}
    \vec{\tau}_b = -\tau_c \frac{\vec{v}_b}{{v_{th}}^q\,|\vec{v}_b|^{1-q}} \,,
\end{equation}
relates the bed-parallel shear stresses, $\vec{\tau}_b$, to the sliding
velocity, $\vec{v}_b$ (Table~\ref{tab:params}).
The yield stress, $\tau_c$,
is modelled using the Mohr--Coulomb criterion,
\begin{equation}
   \tau_c = c_0 + N\,\tan{\phi} \,,
\end{equation}
where cohesion, $c_0$, is assumed to be zero. The friction angle, $\phi$,
varies from 15 to 45\degree as a piecewise-linear function of modern bed
elevation, with the lowest value occuring below modern sea level (0\,m above
sea level, m~a.s.l.) and the highest value occuring above the generalised
elevation of the highest shorelines
\citep[200\,m~a.s.l.,][Fig.~5]{Clague.1981}, thus accounting for
a~weakening of till associated with the presence of marine sediments
(cf. \citealp{Martin.etal.2011}; \citealp[supplement]{Aschwanden.etal.2013};
\citealp{PISM-authors.2015}). Effective pressure, $N$,
is related to the ice overburden stress, $\rho gh$, and the modelled amount of
subglacial water, using a formula derived from laboratory experiments with till
extracted from the base of Ice Stream B in West Antarctica
\citep{Tulaczyk.etal.2000, Bueler.Pelt.2015},
\begin{equation}
    N = \delta \rho gh \, 10^{(e_0/C_c) (1 - (W/W_{max}))} \,,
\end{equation}
where $\delta$ is chosen as 0.02, $e_0$ is a measured reference void ratio and
$C_c$ is a measured compressibility coefficient (Table~\ref{tab:params}). The
amount of water at the base, $W$, varies from zero to $W_{max}=2$\,m, a
threshold above which additional melt water is assumed to drain off
instantaneously. Finally, the bedrock topography responds to ice load
following a bedrock deformation model that includes local isostasy,
elastic lithosphere flexure and viscous astenosphere deformation in an infinite
half-space \citep{Lingle.Clark.1985,Bueler.etal.2007}.
A relatively low viscosity value of $\nu_m = 1\times10^{19}$\,\unit{Pa\,s} is
used for the astenosphere (Table~\ref{tab:params}) in accordance with the
results from regional glacial
isostatic adjustment modelling at the northern Cascadia subduction zone
\citep{James.etal.2009}.

@@

It is generally believed that the Cordilleran ice sheet formed by the
coalescence of several mountain-centred ice caps \citep{Davis.Mathews.1944}.
In our simulations, major ice-dispersal centres, visible on the modelled ice
cover duration maps (Fig.~\ref{fig:duration}), are located over the Coast
Mountains (CM), the Columbia and Rocky mountains (CRM), the Skeena Mountains
(SM), and the Selwyn and Mackenzie mountains (SMKM). The Wrangell and
Saint Elias mountains, heavily glacierized at present, host an ice cap for the
entire length of both simulations, but that ice cap does not appear to be a
major feed to the Cordilleran ice sheet (Fig.~\ref{fig:duration}, WSEM).
Although the Coast, Skeena and Columbia and Rocky mountains (CM, SM, CRM)
are covered by mountain
glaciers for most of the last glacial cycle, providing durable nucleation
centres for an ice sheet, this is not the case for the Selwyn and
Mackenzie mountains (SMKM), where ice cover on the highest peaks is limited to a
small fraction of the last glacial cycle. In other words, the Selwyn and Mackenzie
mountains only appear as a secondary ice-dispersal centre
during the coldest periods of the last glacial cycle. The Northern Rocky
Mountains (Fig.~\ref{fig:duration}, NRM) do not act as a nucleation centre,
but rather as a pinning point for the Cordilleran ice sheet margin coming from
the west.

@@

\begin{figure*}
  \includegraphics{warmbase}
  \caption{Modelled duration of warm-based ice cover during the last
           120\,ka. Long ice cover durations combined with basal
           temperatures at the pressure melting point may explain the strong
           glacial erosional imprint of the Skeena Mountains (SM) landscape.
           Hatches indicate areas that were covered by cold ice only.}
  \label{fig:warmbase}
\end{figure*}

@@

A correlation is observed between the modelled duration of warm based ice cover
(Fig.~\ref{fig:warmbase}) and the degree of glacial modification of the
landscape (mainly in terms of the development of deep glacial valleys and
troughs). We find evidence for this on the slopes of the Coast Mountains, the
the Columbia and Rocky mountains, the Wrangell and Saint Elias mountains, and
radiating off the Skeena Mountains
(Figs.~\ref{fig:duration} and \ref{fig:warmbase};
\citealp[Fig.~2]{Kleman.etal.2010}).
The Skeena Mountains, for example, indeed bear a strong glacial imprint that
indicates ice drainage in a system of distinct glacial troughs emanating in a
radial pattern from the centre of the mountain range
\citep[Fig.~2]{Kleman.etal.2010}. We suggest that
persistent ice cover (Fig.~\ref{fig:duration}) associated with basal ice
temperatures at the pressure melting point (Figs.~\ref{fig:warmbase}
and~\ref{fig:warmfrac}) explains the large-scale glacial erosional imprint on
the landscape. A well-developed network of glacial valleys running to the
north-west on the west slope of the Selwyn
and Mackenzie Mountains (\citealp[Fig.~2]{Kleman.etal.2010}; \citealp[Fig.~8]
{Stroeven.etal.2010}) is modelled to have hosted predominantly warm-based ice
(Fig.~\ref{fig:warmfrac}). However, because it is only glaciated for a
short fraction of the last glacial cycle in our simulations
(Fig.~\ref{fig:duration}), this perhaps indicates that
the observed landscape pattern originates from multiple glacial cycles and
witnesses an increased relative importance of the Selwyn and Mackenzie
mountains ice dispersal centre (Fig.~\ref{fig:duration}, SMKM), prior
to the Late Pleistocene \citep[cf.][]{Ward.etal.2008, Demuro.etal.2012}.

@@

Similarly to other glaciated regions, most glacial traces in the North American
Cordillera relate to the last few millennia of glaciation, because most of the
older evidence has been overprinted by warm-based ice retreat
during the last deglaciation \citep{Kleman.1994, Kleman.etal.2010}. From a
numerical modelling perspective, phases of glacier retreat are more challenging
than phases of growth, because they involve more rapid fluctuations of the ice
margin, increased flow velocities and longitudinal stress gradients, and poorly
understood hydrological processes. The latter are typically included in the
models through simple parametrisations \citep[e.g.][]{Clason.etal.2012,
Clason.etal.2014, Bueler.Pelt.2015}, if included at all. However, next after
the mapping of maximum ice sheet extents during MIS~2 and MIS~4
(Sect.~\ref{sec:mis2} and~\ref{sec:mis4}), geomorphologically-based
reconstructions of patterns of ice sheet retreat during the last
deglaciation provide the second best source of evidence for the validation of
our simulations.

@@

\begin{figure*}
  \includegraphics{deglacshots}
  \caption{Snapshots of modelled surface topography (200\,m contours)
           and surface velocity (colour mapping) during the last deglaciation
           from the GRIP (top panels) and EPICA (bottom panels) 5\,km simulations.
           Dashed segments (A--D) indicate the location of profiles used in
           Figs.~\ref{fig:profiles-grip} and~\ref{fig:profiles-epica}.}
  \label{fig:deglacshots}
\end{figure*}

\begin{figure*}
  \vspace{-1.5mm}  % enough to allow the two figures to fit on one page
  \includegraphics{deglac}
  \caption{Modelled age of the last deglaciation. Areas that have been covered
           only before the last glacial maximum are marked
           in green. Hatches denote re-advance of mountain-centred ice caps and
           the decaying ice sheet between 14 and 10\,ka..
           Dashed segments (A--D) indicate the location of profiles used in
           Figs.~\ref{fig:profiles-grip} and~\ref{fig:profiles-epica}.}
  \label{fig:deglac}
\end{figure*}

@@

Modelled patterns of ice sheet retreat are relatively consistent between the
two simulations (Figs.~\ref{fig:deglacshots} and.~\ref{fig:deglac}). The
southern sector of the modelling domain, including the Puget Lowland, the Coast
Mountains, the Columbia and
and Rocky mountains, and the Interior Plateau of British Columbia, becomes
completely deglaciated by 10\,ka, whereas a significant ice cover remains over
the Skeena, the Selwyn and Mackenzie, and the Wrangell and Saint Elias mountains in the
northern sector of the modelling domain. After 10\,ka, deglaciation continues
to proceed across the Liard Lowland with a radial ice margin retreat towards
the surrounding mountain ranges, consistent with the regional melt water record
of the last deglaciation \citep{Margold.etal.2013}. Remaining ice continues to
decay by retreating towards the Selwyn and MacKenzie, and Skeena mountains.
The last remnants
of the Cordilleran ice sheet finally disappear from the Skeena Mountains at
6.5\,ka (GRIP) and 6.1\,ka (EPICA).

@@

The possibility of late glacial readvances in the North American Cordillera has
been debated for some time \citep{Osborn.Gerloff.1997}, and locally these have
been reconstructed and dated. Radiocarbon-dated end moraines in the Fraser and
Squamish valleys, off the southern tip of the Coast Mountains, indicate
consecutive glacier maxima, or standstills while in overall retreat, one of
which corresponds to the Younger Dryas chronozone \citep{Clague.etal.1997,
Friele.Clague.2002, Friele.Clague.2002a, Kovanen.2002,
Kovanen.Easterbrook.2002}. Although most of these moraines characterise
independent valley glaciers, that may have been disconnected from the waning
Cordilleran ice sheet, the Finlay River area in the Omenica Mountains
(Fig.~\ref{fig:deglac}, OM) presents a different kind of evidence.
There, sharp-crested moraines indicate a late-glacial readvance of local alpine
glaciers and, more importantly, their interaction with larger, lingering
remnants of the main body of the Cordilleran ice sheet in the valleys
\citep{Lakeman.etal.2008}. Additional evidence for late-glacial
alpine glacier readvances includes moraines in the eastern Coast Mountains,
and the Columbia and Rocky mountains \citep{Osborn.Gerloff.1997,
Menounos.etal.2008}.

Although further work is needed to constrain the timing of the late-glacial
readvance, to assess its extents and geographical distribution, and to identify
the potential climatic triggers \citep{Menounos.etal.2008}, it is interesting
to note that the simulation driven by the GRIP record produces a late-glacial
readvance in the Coast Mountains and in the Columbia and Rocky Mountains
(Fig.~\ref{fig:deglac},
left panel). In addition to matching the location of some local readvances, the
GRIP-driven simulation shows that a large remnant of the decaying ice sheet
may still have existed at the time of this late-glacial readvance.
In contrast, the EPICA-driven simulation produces a nearly-continuous
deglaciation with only a tightly restricted late-glacial readvance on the
western slopes of the Saint Elias and the Coast mountains
(Fig.~\ref{fig:deglac}, right panel).

@@

\begin{figure*}
  \includegraphics{lastflow}
  \caption{Modelled deglacial basal ice velocities. Hatches
           indicate areas that remain non-sliding throughout deglaciation
           (22.0--8.0\,ka), notably including parts of the Interior Plateau (IP).
           Note the concentric patterns of deglacial flow in the Liard
           Lowland (LL).
           Sliding grid cells were distinguished from non-sliding grid cells
           using a basal velocity threshold of 1\,\unit{m\,yr^{-1}}.}
  \label{fig:lastflow}
\end{figure*}

@@

On the Interior Plateau of south-central British Columbia, both simulations
produce a retreat of the ice margin towards the north-east
(Fig.~\ref{fig:deglac}), a pattern which is validated by the geomorphological
and stratigraphical record for ancient pro-glacial lakes dammed by the
retreating ice sheet \citep{Perkins.Brennand.2014}. However, the two
simulations differ in the mode of retreat. The GRIP simulation yields an active
retreat with basal sliding towards the ice margin to the south, whereas the
EPICA simulation produces negligible basal sliding on the plateau during deglaciation
(Fig.~\ref{fig:lastflow}). Yet, the Interior Plateau also hosts an impressive
set of glacial lineations which indicate a substantial eastwards flow component
of the Cordilleran ice sheet \citep{Prest.etal.1968, Kleman.etal.2010}.
This Interior Plateau lineation set could therefore present a smoking gun for the
reliability of the presented model results. One explanation for the incongruent
results could be that the missing feedback mechanisms between ice sheet
topography and regional climate resulted in a modelled ice divide of the LGM
ice sheet beingtoo far to the east (Sect.~\ref{sec:mis2}; Fig.~\ref{fig:mis2};
\citealp{Seguinot.etal.2014}). A more westerly-located LGM ice divide would
certainly result in a different deglacial flow pattern over the Interior
Plateau. However, a more westerly-positioned LGM ice divide would certainly be
associated with a thinner ice sheet than that modelled here. Decreased ice
thickness would not promote warm-based conditions but, on the
contrary, enlarge the region of negligible basal sliding
(Fig.~\ref{fig:lastflow}). Thus, a second explanation for the incongruent
results could be that the Interior Plateau lineation system predates
deglaciation ice flow, as perhaps indicated by some eskers
that appear incompatible with these glacial lineations
\citep[Fig.~9]{Margold.etal.2013a}. Finally, a third explanation could be that
local geothermal heat associated with volcanic activity on the Interior Plateau
could have triggered the basal sliding
\citep[cf. Greenland ice sheet;][]{Fahnestock.etal.2001}.

@@

Numerical simulations of the Cordilleran ice sheet through the last glacial
cycle presented in this study consistently produce two glacial maxima during
MIS~4 (61.9--56.5\,ka, 3.8--8.8\,m~s.l.e.) and MIS~2 (23.2--16.8\,ka,
8.40--8.91\,m~s.l.e.), two periods corresponding to documented extensive
glaciations. This result is independent of the palaeo-temperature record used
among the six selected for this study, and thus can be regarded as a robust model
output, which broadly matches geological evidence. However, the timing of the
two glaciation peaks depends sensitively on which climate record is used to
drive the model. The timing of the LGM is best reproduced by the EPICA and
Vostok Antarctic ice core records. It occurs about 2\,ka too early in the
simulation forced by the GRIP ice core record, and occurs even earlier in all other
simulations. The mismatch is largest for the two Northwest Pacific ODP
palaeo-temperature records, which are affected by the weakening of the
California current during the LGM.

In all simulations presented here, ice cover is limited to disjoint mountain
ice caps during most of the glacial cycle. The most persistent nucleation
centres are located in the Coast Mountains, the Columbia and Rocky mountains,
the Selwyn and Mackenzie mountains, and most importantly, in the Skeena
Mountains. Throughout the modelled last glacial cycle, the Skeena Mountains
host an ice cap which appears to be fed by the moisture intruding inland from the
west through a topographic breach in the Coast Mountains. The Skeena ice cap acts
as the main nucleation centre for the glacial reavance towards the LGM
configuration. As indicated
by persistent, warm-based ice in the model, this ice cap perhaps explains
the distinct glacial erosional imprint observed on the landscape of the Skeena
Mountains.

During deglaciation, none of the climate records used can be selected as
producing an optimal agreement between the model results and the geological
evidence. Although the EPICA-driven simulation yields the most realistic timing
of the LGM and, therefore, start of deglaciation, only the GRIP-driven
simulation produces late glacial readvances in areas where these have been
documented. Nonetheless, the patterns of ice sheet retreat are consistent
between the two simulations, and show a rapid deglaciation of the southern
sector of the ice sheet, including a rapid northwards retreat across the Interior
Plateau of central British Columbia. The GRIP-driven simulation then produces a
late-glacial readvance of local ice caps and of the main body of the decaying
Cordilleran ice sheet primarily in the Coast and the Columbia and Rocky Mountains.
In both simulations, this is followed by an
opening of the Liard Lowland, and a final retreat of the remaining ice caps
towards the Selwyn and, finally, the Skeena mountains, which hosts the last
remnant of the ice sheet during the middle Holocene (6.6--6.2\,ka). Our results
identify the Skeena Mountains as a key area to understanding glacial dynamics
of the Cordilleran ice sheet, highlighting the need for further geological
investigation of this region.

@@

% Acknowledgements
\begin{acknowledgements}
Foremost, we would like to thank Shawn Marshall for providing a detailed,
constructive analysis of this study during J.~Seguinot's Ph.D. defence
(September 2014). His comments were used to improve the model set-up.
We are very thankful to Constantine Khroulev, Ed Bueler, and Andy Aschwanden for
providing constant help and development with PISM. This work was supported by the
Swedish Research Council~(VR) grant no.~2008-3449 to A.~P.~Stroeven and by the
German Academic Exchange Service~(DAAD) grant no.~50015537 and a Knut and Alice
Wallenberg Foundation grant to J.~Seguinot. Computer resources were provided
by the Swedish National Infrastructure for Computing (SNIC) allocation
no.~2013/1-159 and~2014/1-159 to A.~P.~Stroeven at the National Supercomputing
Center (NSC).
\end{acknowledgements}
